A parallel runtime framework for communication intensive stream applications

### 2013-03-21 paper.tex
Paper updates

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[conference]{IEEEtran}
%% INFOCOM 2013 addition:
\makeatletter
\def\ps@headings{%
\def\@oddhead{\mbox{}\scriptsize\rightmark \hfil \thepage}%
\def\@evenhead{\scriptsize\thepage \hfil \leftmark\mbox{}}%
\def\@oddfoot{}%
\def\@evenfoot{}}
\makeatother
\pagestyle{headings}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

@@

@@

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{A high-level programming framework to support parallel processing of communication intensive \\ stream applications}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Servesh Muralidharan}
\IEEEauthorblockA{School of Computer Science and\\Statistics\\
Trinity College Dublin\\
Dublin 2, Ireland\\
Email: muralis@tcd.ie}
\and
\IEEEauthorblockN{David Gregg}
\IEEEauthorblockA{School of Computer Science and\\Statistics\\
Trinity College Dublin\\
Dublin 2, Ireland\\
Email: David.Gregg@cs.tcd.ie}}

@@

\begin{abstract}
%\boldmath
Stream applications are often limited in their performance by their underlying communication system. In a typical implementation, the Operating System (OS) handles the majority of network operations. In such cases, the communication stack, which was not designed to handle such tremendous amounts of data, acts as a bottleneck and restricts the performance of the application. Therefore, a framework for computing and communicating in parallel is essential for improving performance.

We propose a high-level programming framework that integrates the packet processing operations with stream applications, and provides a common interface to execute both application and packet processing operations in parallel on multicore processors.

We utilize the ability of the network interface controller (NIC) to classify packets onto multiple hardware queues. This is combined with system software that allows direct access to these queues. We develop an interface in our framework for application processes to read or write in parallel. Our system integrates this with the application by a series of packet processing operations that transform the data from the application to packets suitable for transmission. Representing these tasks as stages in the application, we are able to parallelize them using techniques similar to those used for parallelizing regular computations.

All the communication operations in the framework are performed in userspace, bypassing the OS, to send and receive data directly over the network interface. By overcoming the OS bottleneck, and by parallelizing computations and packet operations together, we are able to boost the performance of stream applications substantially.

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle
%Kevin - vertical spacing
\setlength{\textfloatsep}{15pt}
\section{Motivation}

Stream applications are designed in a manner where they operate on data that flows as a \textit{stream}. This data stream passes through different stages of computations, to be transformed from query to result \cite{Thies:2002:SLS:647478.727935} \cite{Halbwachs91thesynchronous} \cite{Stephens95asurvey}. This differs from a conventional computation model where data is stored and then processed. By eliminating the need to store data, and processing it on the go instead, it is possible for stream applications to process large amounts of data.

Stream applications such as file storage systems, database query management, data mining, etc., operating on vast amounts of data, are often distributed over multiple systems. Traditional stream applications target multicore processors for performing their computations in parallel. As multicore processors grow in power, so do stream applications, processing ever-increasing amounts of data. However, stream applications that operate on large amounts of data at a very high bandwidth can be limited by the communication stack of the OS kernel\cite{Wu:2007:PPB:1323954.1323957}\cite{Wu:2007:PAL:1227865.1228021}\cite{Dobrescu09routebricks:exploiting}\cite{Han:2010:PGS:1851275.1851207}\cite{Kohler2000}. With the continuing improvement of multicore and many-core architectures, it is becoming essential to improve the processing speeds of packet operations along with those of the application in order to scale performance.
\comment{
When performing packet processing operations for communication intensive stream applications, a major challenge is the high levels of performance required. These applications process large numbers of data streams in real time.  If the communication system cannot keep up with the rate at which packets arrive, then they must be dropped or remain unprocessed.
}

In building a framework for such applications on general purpose computers, we can exploit two hardware features that could allow for dramatically improved performance, namely multicore processors and multi queue network controllers. Using these features makes it possible to do much of the packet processing in parallel using the multiple cores commonly available on modern processors.

A major bottleneck for packet processing on general purpose computers is the overhead of the operating system's network stack. In particular, for applications that run as user processes, data must be copied between OS kernel space and user space. Previous research uses alternate mechanisms to access packets from the network hardware directly to overcome this problem\cite{Dobrescu09routebricks:exploiting}\cite{Han:2010:PGS:1851275.1851207}\cite{Kohler2000}. Most of these use modern network hardware in combination with particular system software that allows direct access to hardware-controlled packet queues, dramatically reducing overheads and improving performance. However, constructing stream applications that use these features is difficult. Furthermore, many stream applications have a very specific communication pattern, typically operating on data between a source and sink in the form of a \textit{data stream}. Data in a stream must be processed sequentially, whereas those from separate streams can often be processed in parallel.

Modern network hardware typically has a mechanism for separating arriving packets into multiple queues, where packets that belong to the same flow are\comment{ guaranteed to be} placed in the same queue\cite{micro2008} \cite{intel2010}. We can utilize this feature to operate on multiple streams of data by processing each of the queues in parallel. However, building such applications is complex because it requires explicit parallel programming and interaction with low-level network interfaces. Research focuses on improving the performance of network applications by parallelizing the packet processing tasks on multicore architectures \cite{Dobrescu09routebricks:exploiting} \cite{Han:2010:PGS:1851275.1851207}. We parallelize both the application and packet operations to gain significant improvement in performance.

@@

In this paper we propose a high-level programming framework for building communication intensive stream applications. We focus on integrating the packet processing and application tasks into a single graph and using this to extract data parallelism. The developer describes their program as a graph of \textit{actors}. Each actor consumes one or more \textit{tokens} of data, which may be simple data such as integers, or more complex data such as packets, arrays, or objects and produces one or more tokens. The framework can be used to automatically map the application over multiple parallel cores, and interface with the \texttt{netmap} APIs \cite{Rizzo:2012:RNI:2090147.2103536} for accessing data through the network buffers.

@@

\begin{itemize}
\item A subsystem for interfacing with the network hardware through the \texttt{netmap} APIs to send and receive data from the application in user space without going through the operating system. (Section \ref{ppio})
\item Given a graph that represents the workload to be performed by the application, we construct and integrate the packet processing operations represented as a directed acyclic graph (DAG) to it. (Section \ref{ppe})
\item A run-time system that executes these tasks in parallel and explicitly schedules them on the multicore processor. (Section \ref{ppt})
\end{itemize}

@@

\section{Background}

In this paper, we are concerned with applications that operate on data streams. Example operations of such an application are represented in the stream graph shown in figure \ref{fig1}. This sample application operates by performing encryption on the data that matches predefined rules like file names, access level of user, etc., otherwise it performs compression. Due to the nature of such stream applications, it is possible to distribute the component operations across different CPUs. We are particularly interested in a scenario where groups of networked servers are used to perform these kind of stream operations by distributing them evenly over different nodes.

@@

The main challenge that arises in such a situation is how best to address the intensive computational and communication demands of such applications. Two important features enable us to tackle this challenge. Firstly, the operations of both the application and the network are parallel in nature. Secondly, multicore processors can be used to execute the application.

In our system, we use an efficient parallel processing framework that could be leveraged on multicore processors and multi queue network controllers to achieve efficient communication. The performance of multicore processors continues to dramatically improve the capability of commodity hardware, thus making it possible to meet the computational demand of such applications. In addition, multi queue network interface controllers, designed to handle multiple streams of data, help achieve the necessary communication performance.

An important part of our system is the tasks related to packet processing in protocols such as UDP, IP, etc., We utilize the modular router, Click, for these packet processing tasks. Click is a tool that includes a library of elements which is used in building flexible, re-configurable and modular software routers \cite{Kohler2000}. It offers a high-performance platform that is flexible for rapid development of newer network protocols. An improved version known as SMP Click supports thread-based parallel processing of network related tasks\cite{Chen:2001:FCP:647055.759948}. Until recently, Click's performance in the user space was inferior to its kernel space equivalent. This is mainly due to its extensive use of polling driver in the kernel space that was not possible in user space. Netmap I/O\cite{Rizzo:2012:RNI:2090147.2103536} provides a set of APIs that can directly access the packets from the NIC and was used to improve Click's user space performance. Although other other user space packet-access APIs could be used \cite{Rizzo:2012:RNI:2090147.2103536} \cite{1564468} \cite{Krasnyansky}, the simpler construction of \texttt{netmap} was more suitable to adapt to our framework.

Section \ref{dppl} represents the design principles that we follow to conceptualize our framework. Section \ref{imple} describes the implementation details used in the construction of the system. In section \ref{ppe}, we discuss the challenges in using a system consisting of execution engines to manage the work. The extent to which we extract parallelism ranges, not only to the packet tasks, but also to the low level access of data I/O to the network interface hardware through custom additions to the \texttt{netmap}. Finally in section \ref{app-const}, we present a simple stream application constructed from operations such as compression and encryption in order to evaluate our framework.

@@

\subsection{Userspace Packet I/O}

For a packet processing system to be integrated with an application, it is necessary to have an efficient mechanism for accessing the data from the NIC without interacting with the operating system. Rapid data access here is crucial for the extraction of data parallelism at later stages. \comment{In using the techniques described below we exploit the modern NIC architecture to its full potential.}

@@

The development of virtualization and improved network flow handling has led to hardware level multi queue support in NICs. During transmission, concurrent writes to different queues are possible, enabling multiple processes to send data simultaneously. During reception, the NIC classifies each packet onto one of the receive queues in a technique known as Receive Side Scaling (RSS)\cite{micro2008} \cite{intel2010}. This technique uses the header information or tuples to classify packets onto different hardware queues. Multi queue support can handle several streams of data in parallel and it is exploited extensively in our framework for this reason. By offloading the classification of packets to the NIC, we remove any overhead associated with packet classification in the application. We can also use these hardware queues as a means of balancing the workload across different threads by controlling the number of queues that each process handles. \comment{Since different processes can access these transmit and receive queues, concurrent access to data is possible.}

\subsubsection{Flow based Packet Batching}

Packet batching is a common method used to boost the performance of network applications where several packets are combined and operated upon together, thus reducing the overhead of I/O operations. In our framework, batching is utilized in a form where the number of packets in a batch is varied based on the available space in the hardware queue. This number reflects the amount of data we are able to process before issuing hardware synchronization calls and forms an important means by which we reduce the usage of system calls.

@@

System calls are required in order to synchronize data between the NIC hardware and the buffer memory used for intermediate storage of packets. Previous work describes the overhead associated with per-packet system calls and provides mechanisms for overcoming this overhead\cite{1564468}\cite{Han:2010:PGS:1851275.1851207}\cite{Rizzo:2012:RNI:2090147.2103536}.

@@

\subsection{Packet Processing Graph}

\comment{Understanding the set of packet processing tasks associated with the application is essential to performing them in parallel.} We use a graph-based representation of packet processing tasks to determine the dependencies involved for the different stages and to determine the relation with the different tasks of the application. This graph contains the stages through which the data associated with the application is transformed into network packets and vice versa. It shows which of the tasks can be performed in parallel with those of the application, and highlights those that have a data dependency on a specific task in the application. The graph also enables optimizations based on specific characteristics of the application or the architecture used for execution. Figure \ref{fig2} shows a graph representing User Datagram Protocol (UDP) packet generation.

\subsection{Application Task Graph}

\comment{Application task graph represents the operations associated in performing the computations.} The subset of the application that we are interested in parallelizing is the stream processing component. In such components, the communication flow between tasks is well-defined. Both the packet processing and application tasks can be represent by a directed acyclic graph (DAG) similar to the packet processing graph. The packet processing tasks operate on packets and the application tasks operate on the data. Since we focus more on parallel packet processing, optimizations in relation to the application graph are not discussed. It is assumed that a static version of the application graph is available and we focus more on executing these tasks in parallel with the packet processing tasks.

@@

The sequence of tasks that can be executed in parallel is determined at this stage. The integration stage is used to combine the application graph and the packet processing graph into a larger graph that unifies and represents the overall operations. The dependency relations identified earlier are used in establishing links between the graphs. We determine the sequence of tasks that can be executed in parallel from both the application task graph and packet processing graph. The combined set of tasks is then used by the parallel processing engine.

@@

Two important aspects that the parallel processing engine must address is the scheduling of tasks and balancing the load across the available resources. The parallel processing engine consists of collaborating processes which are bound to specific CPUs and act as engines to execute the workload. Based on their requirements and the available resources, tasks are scheduled across these execution engines. \comment{In this manner, the workload can be distributed flexibly.}

@@

\comment{Most extant stream applications depend on the OS for their communication tasks.}In commodity systems, the interaction with the OS for network operations is a necessary abstraction to handle multiple applications. However, in the case of specialized servers which cater to specific applications, it would be beneficial to interact directly with the network device. Moreover, in the case of intensive network traffic, the OS could act as a bottleneck due to the overhead associated with data flowing through the kernel before being transmitted through the network device\cite{Wu:2007:PPB:1323954.1323957}\cite{Wu:2007:PAL:1227865.1228021}. Eliminating this would be possible, either by executing the application in the kernel space (which could compromise stability of the system), or by using an interface for accessing the NIC buffers from user space. The latter approach is supported by several interfaces such as \texttt{netmap} \cite{Rizzo:2012:RNI:2090147.2103536}, PF\_RING DNA \cite{1564468}, UIO-IXGBE \cite{Krasnyansky}, etc., Even though these interfaces support user space packet access, they lack the ability to be invoked in a efficient manner by the application. More specifically, they do not have any way of optimizing the rate of processing, lacking, for example, the concept of flows for application tasks. The lack of support for interacting with these interfaces in order to read and write data in parallel from concurrent processes or threads adds to the problem. Also, since the application is now responsible for handling network packets, it has to perform the additional operations that would otherwise be carried out by the OS.

@@

Algorithms \ref{s_pack} and \ref{r_pack} use a batch update for sending and receiving packets based on the rate at which data flows through the interface. This is done by maintaining a data structure that has information about the hardware rings, updating it on a periodic basis. Once a fixed amount of space is determined, we update this on our local structure. We then continue to read or write data from the hardware ring, but update the actual limits and position on our local structure. When we have used up the space that was allocated, we update the hardware ring with the changes. In cases where the application is generating sparse traffic, it is possible to frequently update the ring in order to maintain a consistent latency in transmission. Multi queue support is provided by managing the process of selecting the ring to send or receive based on an application determined function. The $f(n)$ present in Algorithm \ref{s_pack} line \ref{s_pack_param} and Algorithm \ref{r_pack} line \ref{r_pack_param} represents this. Choosing a random queue has proven to be effective in our experiments, but if an application requires a specific flow to be maintained, a different function here could provide that support. This would be helpful in cases where the data operations are more predictable, or if the transactions seem to follow a specific pattern. The $threshold$ specified in Algorithm \ref{r_pack} line \ref{r_pack_thres} is determined for each application based on the rate at which it is able to process data. By issuing the system calls such as \textit{ioctl} and \textit{poll} only when required, and effectively handling the multiple queues, we are able to achieve improved performance.

>>>>

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[10pt, conference, compsocconf, reqno]{IEEEtran}

% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

@@

%squeeze

\renewcommand{\arraystretch}{1}

\renewcommand\floatpagefraction{.9}
\renewcommand\topfraction{.9}
\renewcommand\bottomfraction{.9}
\renewcommand\textfraction{.1}
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

\raggedbottom

\addtolength{\floatsep}{-5mm}
\addtolength{\textfloatsep}{-3mm}
%~ \addtolength{\intextsep}{-2mm}
\addtolength{\dbltextfloatsep}{-2mm}
\addtolength{\dblfloatsep}{-1mm}
\addtolength{\abovecaptionskip}{-1mm}
\addtolength{\belowcaptionskip}{-1mm}
\addtolength{\abovedisplayskip}{-3mm}
\addtolength{\belowdisplayskip}{-3mm}
\addtolength{\arraycolsep}{-1mm}

\let\oldthebibliography=\thebibliography
\let\endoldthebibliography=\endthebibliography
\renewenvironment{thebibliography}[1]{%
  \begin{oldthebibliography}{#1}%
    \setlength{\parskip}{0.2ex}%
    \setlength{\itemsep}{0.2ex}%
}%
{%
  \end{oldthebibliography}%
}

\captionsetup{justification=centering}

@@

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{A parallel runtime framework for communication intensive stream applications}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Servesh Muralidharan\IEEEauthorrefmark{2},
Kevin Casey\IEEEauthorrefmark{4}
and
David Gregg\IEEEauthorrefmark{2}}
\IEEEauthorblockA{\IEEEauthorrefmark{2}School of Computer Science and Statistics,
Trinity College Dublin, Dublin 2, Ireland\\
Email: muralis@scss.tcd.ie, David.Gregg.cs.tcd.ie }
\IEEEauthorblockA{\IEEEauthorrefmark{4}DCU\\
Email: }
}
%~
%~ \author{\IEEEauthorblockN{Servesh Muralidharan}
%~ \IEEEauthorblockA{School of Computer Science and\\Statistics\\
%~ Trinity College Dublin\\
%~ Dublin 2, Ireland\\
%~ Email: muralis@tcd.ie}
%~ \and
%~ \IEEEauthorblockN{David Gregg}
%~ \IEEEauthorblockA{School of Computer Science and\\Statistics\\
%~ Trinity College Dublin\\
%~ Dublin 2, Ireland\\
%~ Email: David.Gregg@cs.tcd.ie}}

@@

\begin{abstract}
%\boldmath
Stream applications are often limited in their performance by their underlying communication system. In a typical implementation, the underlying communication relies on operating system to handle majority of the network operations. In such cases, the communication stack, which was not designed to handle such tremendous amounts of data, acts as a bottleneck and restricts the performance of the application.

In this paper, we propose a parallel runtime framework that integrates the communication operations with stream applications, and provides a common parallel processing engine that can execute both the communication and computation operations in parallel on multicore processors.

Our runtime system does this by using a series of communication filters that transform the data from the application to packets suitable for transmission. These communication filters are built using system software that allows direct access to network hardware. We also utilize the ability of the modern network controller to classify packets onto multiple hardware queues by integrating it with the communication filter. Using this our runtime framework is able to construct a unified stream graph based on the application graph provided. This when used with a static schedule based on a given architecture, can process the data streams in parallel.

We are able to parallelize the stream application and achieve speedups of more than a factor of eight in all the applications we tested. The results show that our system scales to as many as 24 parallel processes on a multicore computer, and achieves speedups of more than a factor of ten in some cases compared to sequential implementations.

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle
%Kevin - vertical spacing % Servesh - Fixed it on global command
%\setlength{\textfloatsep}{15pt}
\section{Motivation}

Applications that follow streaming model~\cite{Thies:2002:SLS:647478.727935,Halbwachs91thesynchronous,Stephens95asurvey,muthukrishnan2005data} are expressed as a set of filters that operate on one or more tokens of data. The flow of these data tokens through the filters is referred to as a \textit{stream}. Since the data is accessed sequentially in the form of a stream the memory requirement to process the data is considerably less compared to the size of the enitre stream, thereby making it possible to operate on very large data sizes. Typical applications based on this model operate on data ranges in the order of terabytes or even bigger. Applications such as financial trading systems~\cite{agarwal2009faster}, database systems~\cite{Golab03datastream}, network data analysis~\cite{Gilbert01quicksand:quick,Sullivan:1998:TSM:1268256.1268258} are few examples that follow the streaming model and operate on very large data sets.

Traditional stream applications target multicore processors for performing their computations in parallel. As multicore processors grow in size, so do stream applications, processing more streams of data in parallel. However, in such cases the applications are typically limited by the traditional communication systems~\cite{Wu:2007:PPB:1323954.1323957,Wu:2007:PAL:1227865.1228021,Dobrescu09routebricks:exploiting,Han:2010:PGS:1851275.1851207,Kohler2000}. The reason being these systems have limited support for operating on parallel streams and are burdened by abstraction layers that provide compatibility for non-streaming applications.

Several high speed networks~\cite{fredj2011survey} exists that could handle such communication requirements but these require expensive hardware and specialized programming skills. Commodity network adapters based on 10-gigabit ethernet (10GbE) are comparatively cheaper alternative and are being widely deployed in clusters and grids in data centers. Previous research\cite{Romanow03anoverview,Feng:2003:OEN:1048935.1050200} have been done on optimizing ethernet for improving communication in such cases. However, they do not use many of the features provided by modern day network adapters such as hardware queues, flow control, etc., Also, they do not focus specifically on streaming applications and hence making it difficult to utilize them properly.

@@

In this paper we propose a parallel processing framework for building stream applications that are bound by very large communication. A simplistic framework where the filters are described as \textit{actors} and the application is represented by connecting such \textit{actors} into a directed acyclic graph (DAG). Then our framework is used to establish, (1)the communication system that allows these \textit{actors} to send and receive data across each other and (2)the computation system that executes these tasks in parallel on a given multicore system. The main contributions can be described as,

\begin{enumerate}
\item Parallel multi point communication layer, which provides the ability for stream actors to communication across the network.
\item Execution system that can run the given data parallel filters on the multicore system
\item Utilizes network hardware features such as queues and flow control to support stream applications

\end{enumerate}

\comment{
We focus on integrating the packet processing and application tasks into a single graph and using this to extract data parallelism. The developer describes their program as a graph of \textit{filters}. Each actor consumes one or more \textit{tokens} of data, which may be simple data such as integers, or more complex data such as packets, arrays, or objects and produces one or more tokens. The framework can be used to automatically map the application over multiple parallel cores, and interface with the \texttt{netmap} APIs \cite{Rizzo:2012:RNI:2090147.2103536} for accessing data through the network buffers.

@@

\begin{itemize}
\item A subsystem for interfacing with the network hardware through the \texttt{netmap} APIs to send and receive data from the application in user space without going through the operating system. (Section \ref{ppio})
\item Given a graph that represents the workload to be performed by the application, we construct and integrate the packet processing operations represented as a directed acyclic graph (DAG) to it. (Section \ref{ppe})
\item A run-time system that executes these tasks in parallel and explicitly schedules them on the multicore processor. (Section \ref{ppt})
\end{itemize}
}

@@

\comment{
With the continuing improvement of multicore and many-core architectures, it is becoming essential to improve the processing speeds of packet operations along with those of the application in order to scale performance.
}
\comment{
When performing packet processing operations for communication intensive stream applications, a major challenge is the high levels of performance required. These applications process large numbers of data streams in real time.  If the communication system cannot keep up with the rate at which packets arrive, then they must be dropped or remain unprocessed.
}

\section{Introduction}

Our main objective is to optimize performance of streaming applications by First, providing a speaclized communication link between the filters and Second, extracting data parallel actors and executing them in parallel. Stream applications usually consists of stateless and stateful filters. Stateful filters are those that are depedendent on the previous execution and hence making it difficult to replicate them. Stateless filters do not pose such limitations and can be freely replicated based on the available data streams. In this paper, we focus on extracting data parallelism based on stateless filters and streams. Example operations of such an application are represented in the stream graph shown in Figure \ref{fig1}. All the filters used in this application are essentially stateless filters. This sample application operates by performing encryption on the data that matches predefined rules like file names, access level of user, etc., otherwise it performs compression. Due to the nature of such stream applications, it is possible to replicate and distribute the filters across different CPUs. We are particularly interested in a scenario where groups of networked servers are used to perform these kind of stream operations by distributing data streams over multiple nodes.

@@

The main challenge that arises in such a situation is how best to address the intensive computational and communication demands of such applications. Two important features enable us to tackle this challenge in our framework. Firstly, the operations of both the application and the network are designed to be stateless filters. Secondly, communication techniques are employed that would enable multicore processors to perform operations related to computation and communication in parallel.

In building a framework for such applications on general purpose computers, we can exploit two hardware features that could allow for dramatically improved performance, namely multicore processors and multi queue network controllers. Using the multiple queues it makes it possible to do much of the communication in parallel and using the multiple cores the filters from the stream application can be executed concurrently.

A major bottleneck for communication on general purpose computers is the overhead of the operating system's network stack. In particular, for applications that run as user processes, data must be copied between OS kernel space and user space. Previous research uses alternate mechanisms to access data from the network hardware directly to overcome this problem\cite{Dobrescu09routebricks:exploiting}\cite{Han:2010:PGS:1851275.1851207}\cite{Kohler2000}. Most of these use modern network hardware in combination with specialized system software that allows direct access to hardware-controlled packet queues, dramatically reducing overheads and improving performance. However, constructing stream applications that use these features is difficult. Furthermore, many stream applications have a very specific communication pattern, typically operating on a \textit{flow}, a specific \textit{stream} between two filters. Data in a \textit{stream} must be processed sequentially, whereas different \textit{flows} can often be processed in parallel.

Newer network hardware has introduced a mechanism for separating arriving data into multiple queues, where data that belong to the same \textit{flow} are\comment{ guaranteed to be} placed in the same queue\cite{micro2008} \cite{intel2010}. We can utilize this feature to operate on multiple streams of data by processing each of the queues in parallel. However, building applications that can use this feature directly is a challenging task because it requires explicit parallel programming and interaction with low-level network interfaces. Previous research~\cite{Dobrescu09routebricks:exploiting,Han:2010:PGS:1851275.1851207} that focuses on improving the performance of network applications in such a manner shows the complexity involved in building such a system.

\comment{We parallelize both the application and packet operations to gain significant improvement in performance.}

In our framework, we use a more elegant approach. We define a runtime framework that already consists of the communication filters that are pre-built. So inorder to implement a stream application such as that shown in Figure~\ref{fig1}, all we need to do is define the computation filters and connect them in the framework to the pre-built communication filters. On execution, our runtime framework is able to replicate the stateless filters and the communication filters on to the different multiple cores, thereby executing the operations belonging to different \textit{flows} in parallel.

\comment{
 parallel processing framework that could be leveraged on multicore processors and multi queue network controllers to achieve efficient communication. The performance of multicore processors continues to dramatically improve the capability of commodity hardware, thus making it possible to meet the computational demand of such applications. In addition, multi queue network interface controllers, designed to handle multiple streams of data, help achieve the necessary communication performance.

An important part of our system is the tasks related to packet processing in protocols such as UDP, IP, etc., We utilize the modular router, Click, for these packet processing tasks. Click is a tool that includes a library of elements which is used in building flexible, re-configurable and modular software routers \cite{Kohler2000}. It offers a high-performance platform that is flexible for rapid development of newer network protocols. An improved version known as SMP Click supports thread-based parallel processing of network related tasks\cite{Chen:2001:FCP:647055.759948}. Until recently, Click's performance in the user space was inferior to its kernel space equivalent. This is mainly due to its extensive use of polling driver in the kernel space that was not possible in user space. Netmap I/O\cite{Rizzo:2012:RNI:2090147.2103536} provides a set of APIs that can directly access the packets from the NIC and was used to improve Click's user space performance. Although other other user space packet-access APIs could be used \cite{Rizzo:2012:RNI:2090147.2103536} \cite{1564468} \cite{Krasnyansky}, the simpler construction of \texttt{netmap} was more suitable to adapt to our framework.
}

Section \ref{dppl} represents the design principles that we follow to conceptualize our framework. Section \ref{imple} describes the implementation details used in the construction of the system. In section \ref{ppe}, we discuss the challenges in using a system consisting of execution engines to manage the work. \comment{The extent to which we extract parallelism ranges, not only to the packet tasks, but also to the low level access of data I/O to the network interface hardware through custom additions to the \texttt{netmap}.} Finally in section \ref{app-const}, we present a simple stream application constructed from operations such as compression and encryption and evaluate our framework.

@@

\subsection{Communication filters}

\comment{Understanding the set of packet processing tasks associated with the application is essential to performing them in parallel.} We use a graph-based representation of filters related to packet processing to determine the dependencies involved for the different stages and to determine the relation with the different tasks of the application. This graph contains the stages through which the data associated with the application is transformed into network packets and vice versa. It shows which of the tasks can be performed in parallel with those of the application, and highlights those that have a data dependency on a specific task in the application. The graph also enables optimizations based on specific characteristics of the application or the architecture used for execution. Figure \ref{fig2} shows a graph representing User Datagram Protocol (UDP) packet generation.

\subsection{Userspace communication}

For a communication system to be integrated with an application, it is necessary to have an efficient mechanism for reading and writing packets to and from the NIC without interacting with the operating system. Rapid data access here is crucial for the extraction of data parallelism at later stages. \comment{In using the techniques described below we exploit the modern NIC architecture to its full potential.}

@@

The development of virtualization and improved network flow handling has led to hardware level multi queue support in NICs. During transmission, concurrent writes to different queues are possible, enabling multiple processes to send data simultaneously. During reception, the NIC classifies each packet onto one of the receive queues in a technique known as Receive Side Scaling (RSS)\cite{micro2008} \cite{intel2010}. This technique uses the header information or tuples to classify packets onto different hardware queues. Multi queue support can handle several streams of data in parallel and it is exploited extensively in our framework for this reason. By offloading the classification of packets to the NIC, we remove any overhead associated with packet classification in the application. We can also use these hardware queues as a means of balancing the workload across different threads by controlling the number of queues that each thread handles. \comment{Since different processes can access these transmit and receive queues, concurrent access to data is possible.}

\subsubsection{Flow based communication}

Combining multiple packets that belong to a flow known as \texttt{batching} is a common method used to boost the performance of network applications where several packets are combined and operated upon together, thus reducing the overhead of I/O operations. In our framework, batching is utilized in a form where the number of packets in a batch is varied based on the available space in the hardware queue. This number reflects the amount of data we are able to process before issuing hardware synchronization calls and forms an important means by which we reduce the usage of system calls.

@@

System calls are required in order to synchronize data between the NIC hardware and the buffer memory used for intermediate storage of data. Previous work describes the overhead associated with per-packet system calls and provides mechanisms for overcoming this overhead\cite{1564468}\cite{Han:2010:PGS:1851275.1851207}\cite{Rizzo:2012:RNI:2090147.2103536}.

@@

\subsection{Computation filters}

\comment{Computation filters represents the operations associated . The subset of the application that we are interested in parallelizing is the stream processing component.} In stream processing model, the communication flow between tasks is well-defined. \comment{This is usually represented by a synchronous data flow graph (SDF).} The filters associated with the computation can be connected to the communication filters in the same directed acyclic graph (DAG) used to represent the application. \comment{The packet processing tasks operate on packets and the application tasks operate on the data. Since we focus more on parallel packet processing, optimizations in relation to the application graph are not discussed. It is assumed that a static version of the application graph is available and we focus more on executing these tasks in parallel with the packet processing tasks.} The computation filters are designed by the user and are programmed into the runtime framework which already consists of the communication filters.

@@

\comment{The sequence of tasks that can be executed in parallel is determined at this stage. The integration stage is used to combine the application graph and the packet processing graph into a larger graph that unifies and represents the overall operations. The dependency relations identified earlier are used in establishing links between the graphs. We determine the sequence of tasks that can be executed in parallel from both the application task graph and packet processing graph. The combined set of tasks is then used by the parallel processing engine.}

This stage essentially connects the corresponding communication filter to the computation filter defined earlier. Based on the stream graph that is provided, the necessary communication filters are added. This also determines the unique \textit{flows} that exists between the filters and is used later to replicate and execute them in parallel.

@@

Two important aspects that the parallel processing engine must address is the scheduling of filters and balancing the load across the available resources. The parallel processing engine consists of collaborating processes which are bound to specific CPUs and act as engines to execute the workload. Based on their requirements and the available resources, filters are scheduled across these execution engines. We assume a static schedule of the filters on to the given multiple cores is provided for execution. \comment{In this manner, the workload can be distributed flexibly.}

@@

\comment{Most extant stream applications depend on the OS for their communication tasks.}In commodity systems, the interaction with the OS for network operations is a necessary abstraction to handle multiple applications. However, in the case of specialized servers which cater to specific applications, it would be beneficial to interact directly with the network device. Moreover, in the case of intensive network traffic, the OS could act as a bottleneck due to the overhead associated with data flowing through the kernel before being transmitted through the network device\cite{Wu:2007:PPB:1323954.1323957}\cite{Wu:2007:PAL:1227865.1228021}. Eliminating this would be possible, either by executing the application in the kernel space (which could compromise stability of the system), or by using an interface for accessing the NIC buffers from user space. The latter approach is supported by several interfaces such as \texttt{netmap} \cite{Rizzo:2012:RNI:2090147.2103536}, PF\_RING DNA \cite{1564468}, UIO-IXGBE \cite{Krasnyansky}, etc., Even though these interfaces support user space packet access, they lack the ability to be invoked in a efficient manner by the application. More specifically, they do not have any way of optimizing the rate of processing, lacking, for example, the concept of flows for application tasks. The application tasks are essentially the computation filters defined in our design principles. The lack of support for interacting with these interfaces in order to read and write data in parallel from concurrent processes or threads adds to the problem. Also, since the application is now responsible for handling network packets, it has to perform the additional operations that would otherwise be carried out by the OS.

@@

Algorithms \ref{s_pack} and \ref{r_pack} use a batch update for sending and receiving packets based on the rate at which data flows through the interface. This is done by maintaining a data structure that has information about the hardware rings, updating it on a periodic basis. Once a fixed amount of space is determined, we update this on our local structure. We then continue to read or write data from the hardware ring, but update the actual limits and position on our local structure. When we have used up the space that was allocated, we update the hardware ring with the changes. In cases where the application is generating sparse traffic, it is possible to frequently update the ring in order to maintain a consistent latency in transmission. Multi queue support is provided by managing the process of selecting the ring to send or receive based on an application determined function. The $f(n)$ present in Algorithm \ref{s_pack} line \ref{s_pack_param} and Algorithm \ref{r_pack} line \ref{r_pack_param} represents this. Choosing a random queue ($f(n)=X \sim U([0,n])$ n=no of queues)has proven to be effective in our experiments, but if an application requires a specific flow to be maintained, a different function here could provide that support. This would be helpful in cases where the data operations are more predictable, or if the transactions seem to follow a specific pattern. The $threshold$ specified in Algorithm \ref{r_pack} line \ref{r_pack_thres} is determined for each application based on the rate at which it is able to process data. By issuing the system calls such as \textit{ioctl} and \textit{poll} only when required, and effectively handling the multiple queues, we are able to achieve improved performance.
